Human Factors Influencing Visual Statistical Inference

### 2015-08-28 sociology_paper.Rnw
read through Data Collection

Figure~\ref{fig:elect} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are showing  data generated under a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 out of 72 observers picked the  plot of the observed data, located in panel number $4^2-3$ in the lineup \footnote{The little piece of calculus used for describing the location of the data plot in the lineup provides a small cognitive hurdle that is supposed to enable the interested reader to inspect the lineup without being biased by knowing the location of the data plot.}.  The corresponding $p$-value is 0.0077, indicating sufficient evidence to reject the null hypothesis. 
% , with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).

How do we interpret this finding, though? For that, we need to know the context of the data and we need to have more information about how the null plots were generated. 
This particular example investigates the results from the 2012 US presidential election in comparison to  poll results just prior to the election. 
Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol. The data consists of the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect} shows an `electoral building' \citep{mosley2010} where each state in the union is represented by a rectangle. The difference in poll results is plotted horizontally, while the height of each box corresponds to the number of that state's electoral votes. Color indicates party affiliation.  The null hypothesis is that ``the election results were consistent with the polls''. Therefore the polling results provide the null model from which data is simulated. A normal model with mean and standard deviation based on a poll and its margin of error is used to simulate different scenarios that might have resulted on election day, if the polls were on target. Each null data set is generated as a set of draws from this model. % normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. 
These samples are plotted as electoral buildings, and the plot based on the actual election results is placed randomly among the null plots in a lineup of size $5\times 4$. If the null hypothesis is true, the  plot of the actual data should look  like any of the other plots, and not be identifiable by an observer. Figure~\ref{fig:tower} shows the plot of the electoral building with the additional context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

\begin{figure}[hbtp] 
   \centering
   \begin{minipage}[c]{0.49\textwidth}
   \includegraphics[width=\textwidth]{tower.pdf}
   \end{minipage}
    \hfill
   \begin{minipage}[c]{0.45\textwidth}
   Poll aggregator 1 \\
      \includegraphics[width=\textwidth]{freedom-sim.pdf} 
   Poll aggregator 2\\
      \includegraphics[width=\textwidth]{rcp-sim.pdf} 
   \end{minipage}
   \caption{ \label{fig:tower} Electoral building plot of the results of the 2012 U.S. Presidential Election (left). On the right two histograms showing simulated poll results for the Democratic presidential candidate using  polling averages collected at two different sources, Freedom's Lighthouse Averages (top), RealClearPolitics (bottom).  If the polls matched the electoral result the simulated results would be centered at the electoral result. This happens for the RealClearPolitics polls (bottom) but not for Freedom's Lighthouse.  For the latter only 5.33\% of simulated results were as high as the actual election result, which converts to a $p$-value of 0.0533, almost a statistically significant difference. It hints at the polls from this source under-estimating the vote for Obama. This source was used to generate the null plots in the lineup in Figure~\ref{fig:elect}. The $p$-value from the lineup protocol, was 0.0077, which more strongly concludes that these polls mismatched the actual result.}
\end{figure}

@@

A lineup can be evaluated by a single person or multiple observers. A distribution similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario, is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis.  To avoid expectation errors \citep{meilgaard}, particularly in emotionally charged areas such as election results, observers should not be aware of the data that constitutes a lineup, and should therefore not have seen the actual data plot before inspecting the lineup. This is the reason that we explained the context of the election problem only after you had a chance to evaluate the lineup.%the election example, above, the scenario was explained after the lineup question, in the text.

When presenting a lineup to an observer  the accompanying question should be phrased in as general a manner as possible, 
%The question that is asked of the observer should be as general as possible, 
effectively asking the observer to pick the plot that is most different, and allowing observers to provide their own reasons for their choice. This ensures all possible deviations from the null hypothesis to be detected. 
However, if lineup tests are run in head-to-head comparisons with conventional statistical tests, such as the experiments  in  \cite{majumder:2013} or \cite{tengfei:2013}, these questions have to be phrased much more specifically, in order to avoid errors of type III \citep{mosteller:48} consisting of correctly rejecting the null hypothesis for a wrong reason. 
%In some of the studies, the ones described in  \cite{majumder:2013} very specific questions were asked, because the experiments were being conducted to compare results from the lineup protocol with those of conventional tests. 
In those experiments, the structure in the data is strictly controlled in the simulation process, which allows for specific questions to be asked. 
In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is the difference between the two heights. 
Most observers based their decision on this -- besides  `asymmetry' between the towers (a reason suggested by us), most free-form responses for the reason of choice describe a comparison between the red and blue towers. 
In the responses for this lineup we see something that is very typical of responses in lineups: it is only a few panels from the lineup that observers pick, while most panels do not get picked or get picked only a few times. Out of a total of 72~responses, 23 observers picked panel~\#8, 12 observers picked the (data) panel \#13, 10 observers picked panel~\#4, 6 observers picked panel \#7, and all of the remaining panels were picked at most three times. From the panels picked, we see that the top three choices  are all extreme with respect to heights of blue versus red towers:  both panels \#8 and \#13 show a large difference between the height of the towers, while panel \#4 is the only panel, in which the red tower is higher than the blue one. 
% results from the turk study:
% panel:   1 10 11 12 13 14 15 16 18 19  2 20  3   4  5  6  7  8  9 
% #picks:  2  1    0  1 12    3  3    1  1   1  2   1  0 10  2  2  6 23  1

@@

We have conducted a series of experiments (see Table~\ref{tbl:visual_stat} for an overview)
%There have been ten experiments conducted 
using Amazon's Mechanical Turk~\citep{turk} (MTurk) for a variety to purposes: validating the protocol against existing tests, comparing plot designs, and evaluating structure in data analysis problems. In all of the experiments we collected demographic details, such as gender, age bracket, and educational background.  Mining this data will provide a basis for evaluating
%are being used to evaluate 
the effects of different covariates on an individual's ability to evaluate lineups. The design of some of the experimets allow a further investigation into aspects beyond participants' demographics: experiments~5, 6 and 7 can be used to investigate short term learning trends, while experiment~9 allows insight on whether the  positioning of the  data plot in the lineup has an effect on detection rate. 
 %Each of these experiments collected demographic details of the subjects. 
Section~\ref{sec:factor_performance} discusses human factors that we expect to be influential on the performance of the observer. Section~\ref{sec:exp_design} describes the methods used to assess the effects, and Section~\ref{sec:result_socio} describes the results on what we have learned about human factors affecting lineup evaluation.

@@

%\subsection{Signal in the Data}
\item{\bf Signal in the Data:}
A visual test statistic is chosen for a particular task. If it is well-designed then departures from the null hypothesis will be visible in the plot, especially if the signal is strong. An important factor that helps an observer to identify the actual data plot in a lineup is the strength of the signal in the data. Simulated data was generated for all of the experiments, except exps.~4 and~9, enabling the study of signal strength and actual data plot detection. In experiment~4 we took a different approach by varying the sampling size, which also allowed us to investigate the effect a design had on detection rate while varying signal/noise ratio. Primarily the results were as expected, that as the signal strength increased observers more frequently picked the actual data plot as different from the null plots.

@@

\begin{enumerate} \leftmargin 5cm  \itemsep 0in
\item {\it Age group}, with categories set to be 18-25, 26-30, 31-35, 36-40, 41-45, 46-50, 51-55, 56-60, above 60.
\item {\it Gender}, male or female.
\item {\it Education level}, with levels being high school or less, some undergraduate courses, undergraduate degree, some graduate courses, and graduate degree.
\item {\it Geographical location}, collected from the IP address of the participants' computer, as latitude, longitude, city and country. 
\end{enumerate}
%
%Instead of collecting exact age, 9 levels of age were collected. They are 18-25, 26-30, 31-35, 36-40, 41-45, 46-50, 51-55, 56-60, above 60. There were five levels of academic information. They are (1) High school or less , (2) Some under grad course, (3) Under graduate degree, (4) Some graduate courses,  (5) Graduate degree.  Geographical locations were collected using the true public ip addresses of the participants' computer. This provides latitude and longitude of the ip address as well as the city and country information. 
%
Let $Y_{ij}$ denote the response from observer $i$ on a lineup $j$, with $Y_{ij} =1$ if the actual data plot is chosen, otherwise $Y_{ij} =0$. The factors are examined in association with the observer's response using a logistic regression with random effect terms:
\begin{equation} \label{eqn:demographic_response}
g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
\end{equation}
where $\pi_{ij}=  E(Y_{ij})$ is the probability that  observer $i$ picks the actual data plot from lineup $j$, %We model this in a generalized mixed effects model of the form
%\begin{equation} \label{eqn:demographic_response}
%g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
%\end{equation}
 $\mu$ is an overall population average, %probability for picking out the data plot from a lineup HH: mu is not a probability, it is on the scale of the link function
 $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are the effects of age group $k(i)$, gender $l(i)$, education level $m(i)$ and country name $s(i)$, respectively for observer $i$. The term $\ell_j$ is a random intercept predicting lineup difficulty level and we assume independence and normality of the errors, i.e.\ $\ell_j \sim N(0, \sigma_\ell^2)$.  $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.

@@

It should be noted that we are examining the observer's performance, when we model response as detected or not, but this is not the goal of visual inference. Visual inference is constructed to measure the significance of structure discovered in data. It is expected that some observers will be more skilled at reading data plots, and hence, more readily detect the plot that is different. It is also expected that as observer's gain experience in evaluating lineups that they become more proficient in reading data plots, particularly if feedback is given on whether the actual data plot was chosen or not. Choosing the actual data plot will be more difficult in some lineups than others, and indeed should happen purely by chance in some lineups.%, so ``correct'' is more of a convenient rather than ideal word to use for this analysis. 
So in this context, detected, or not, is used as a response to examine individual differences.

@@

%Multiple lineups were sequentially shown to each observer for evaluation. 
Let $Y_{ijk}$ denote the response from observer $i$ on lineup $j$ at their $k$th evaluation attempt, where $Y_{ijk}=1$ if the observer detected the actual data plot otherwise  $Y_{ijk}=0$. Let $\pi_{ijk}=  E(Y_{ijk})$ be the probability that  observer $i$ picks the actual data plot from lineup $j$ in their $k$th attempt. Learning trend is assessed using a generalized mixed effects model of the form
\begin{equation} \label{eqn:trend_response}
g( \pi_{ijk} )= \mu + \alpha_k + u_i +  a_{i} k + \ell_j,  
\end{equation}
where $\mu$ is an overall population average, %probability for picking out the data plot from a lineup, 
$\alpha_k$ is the effect of the $k$th attempt on the probability, with $\alpha_1 = 0$ and $k = 1, ..., K$, $u_i$ and $a_i$ are observer specific random effects, $i = 1, ..., I$. The term, $u_i$ is a random intercept, describing a basic subject-specific ability, with $u_i \sim N(0, \sigma_u^2)$. 
The term $a_i$ is a random slope capturing an individual's specific learning effect over the course of $K$ attempts, where $a_i \sim N(0, \sigma_a^2)$. 
For $\ell_j$ a normal distribution, $N(0, \sigma_\ell^2)$, is assumed, and $\ell_j$ is a random intercept predicting lineup difficulty level. $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$. The inverse link function, $g^{-1}(.)$, from equation~\ref{eqn:trend_response} leads to the estimate of the subject and the lineup specific probability of successful evaluation in the $k$th attempt by a single observer as 
\begin{equation} \label{eqn:trend_power}
\hat p_{ijk} =  g^{-1}(\hat{\mu} + \hat{\alpha}_k + \hat{u}_i +  \hat{a}_i k + \hat{\ell}_j).
\end{equation}

When time taken to evaluate a lineup is used as the response, let $Z_{ijk}$ denote the logarithm of time taken for an observer $i$ to evaluate a  lineup $j$ in his/her $k$th attempt. Let $\mu_{ijk}=  E(Z_{ijk})$ be the average of log(time taken) by  observer $i$ to pick the data panel from lineup $j$ in his/her $k$th attempt. We model this in a mixed effects model of the form
%\hh{
\begin{equation} \label{eqn:trend_time}
Z_{ijk} = \mu + \alpha_1 + \alpha k + u_i +  a_{i} k + \ell_j + \epsilon_{ijk},  
\end{equation}
where $\mu$ represents overall average of log time taken by an observer to evaluate a lineup, $\alpha$ is the average change in log time taken for each additional attempt,  $\alpha_1$ is an offset in log time taken for the first attempt. All other effects are random effects: $u_i$ is a subject-specific intercept representing individual speed of an observer with $u_i \sim N(0, \sigma_u^2)$, $a_i$ is a subject-specific slope representing the deviation of the speed-up (or -down) by attempt $k$, with $a_i \sim N(0, \sigma_a^2)$, $\ell_j$ is a lineup-specific random effect for the time needed to evaluate a lineup, $\ell_j \sim N(0, \sigma_\ell^2)$ and the overall error $\epsilon_{ijk} \sim N(0, \sigma^2)$.
Equation~\ref{eqn:trend_time} leads to the estimate of the subject and the lineup specific time taken for an evaluation in $k$th attempt by a single observer as 
\begin{equation} \label{eqn:trend_time_est}
\hat \mu_{ijk} =  \hat{\mu} + \hat{\alpha_1}+ \hat{\alpha}k + \hat{u}_i +  \hat{a}_i k + \hat{\ell}_j.
\end{equation}

To fit all these mixed effect models the function {\tt lmer()} is used from R package {\tt lme4} by \cite{lme4:2015, lme4:paper}. To obtain the $p$-values of fixed effect parameters estimates, normal approximation is used to compute $Z$ scores  as the ratio of estimates to  estimated standard errors.

>>>>

Figure~\ref{fig:elect} displays an example lineup, one of the plots is based on the observed data, while the remaining 19 plots show  data generated from a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 out of 72 observers picked the  data plot located in panel number $4^2-3$ in the lineup\footnote{The little piece of calculus used for describing the location of the data plot in the lineup provides a small cognitive hurdle that is supposed to enable the interested reader to inspect the lineup without being biased by knowing the location of the data plot.}.  The corresponding $p$-value is 0.0077, indicating sufficient evidence to reject the null hypothesis. 
% , with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).

How do we interpret this finding, though? For that, we need to know the context of the data and we need to have more information about how the null plots were generated. 
This particular example investigates the results from the 2012 US presidential election in comparison to  poll results just prior to the election. 
Although this example is more simplistic than most of the tests conducted to date, it serves the purpose of illustrating the lineup protocol. The data consists of the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect} shows an `electoral building' \citep{mosley2010} where each state in the union is represented by a rectangle. The difference in poll results is plotted horizontally, while the height of each box corresponds to the number of that state's electoral votes. Color indicates party affiliation.  The null hypothesis is that ``the election results are consistent with the polls''. Therefore the polling results provide the null model from which data is simulated. A normal model with mean and standard deviation based on a poll and its margin of error is used to simulate different scenarios that might have resulted on election day, if the polls were on target. Each null data set is generated as a set of draws from this model. % normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. 
These samples are plotted as electoral buildings, and the plot based on the actual election results is placed randomly among the null plots in a lineup of size $5\times 4$. If the null hypothesis is true, the  data plot should look  like any of the other plots, and not be identifiable by an observer. Figure~\ref{fig:tower} shows the plot of the electoral building with the additional context information and labels.
 %A party needs 270 electoral votes or more to win the presidential election. 
%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=0.8\linewidth]{electoral-5-13.pdf} 
%   \caption{Which plot looks the most different from the other plots?  Lineup \#1 - electoral building; another set of four lineups can be found in figs.~{fig:elect.2} through~{fig:elect.5}}
%   \label{fig:elect.1}
%\end{figure}

\begin{figure}[hbtp] 
   \centering
   \begin{minipage}[c]{0.49\textwidth}
   \includegraphics[width=\textwidth]{tower.pdf}
   \end{minipage}
    \hfill
   \begin{minipage}[c]{0.45\textwidth}
   Poll aggregator 1 \\
      \includegraphics[width=\textwidth]{freedom-sim.pdf} 
   Poll aggregator 2\\
      \includegraphics[width=\textwidth]{rcp-sim.pdf} 
   \end{minipage}
   \caption{ \label{fig:tower} Electoral building plot of the results of the 2012 U.S. Presidential Election (left). On the right two histograms showing simulated poll results for the Democratic presidential candidate using  polling averages collected from two different sources, Freedom's Lighthouse Averages (top), RealClearPolitics (bottom).  If the polls matched the electoral result the simulated results would be centered at the electoral result. This happens for the RealClearPolitics polls (bottom) but not for Freedom's Lighthouse.  For the latter, only 5.33\% of simulated results were as high as the actual election result, which converts to a $p$-value of 0.0533, indicating an almost statistically significant difference. It hints at the polls from this source under-estimating the vote for Obama. This source was used to generate the null plots in the lineup in Figure~\ref{fig:elect}. The $p$-value from the lineup protocol, was 0.0077, which more strongly concludes that these polls mismatched the actual result.}
\end{figure}

@@

A lineup can be evaluated by a single person or multiple observers. A distribution similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario, is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis.  To avoid expectation errors \citep{meilgaard}, particularly in emotionally charged areas such as election results, observers should not be aware of the data that constitutes a lineup, and should therefore not have seen the actual data plot before inspecting the lineup. This is the reason that we revealed the context of the election problem only after the interested reader had a chance to evaluate the lineup.%the election example, above, the scenario was explained after the lineup question, in the text.

When presenting a lineup to an observer  the accompanying question should be phrased in as general a manner as possible, 
%The question that is asked of the observer should be as general as possible, 
effectively asking the observer to pick the plot that is most different, and allowing observers to provide their own reasons for their choice. This ensures all possible deviations from the null hypothesis to be detected. 
However, if lineup tests are run in head-to-head comparisons with conventional statistical tests, such as the experiments  in  \cite{majumder:2013} or \cite{tengfei:2013}, these questions have to be phrased much more specifically, in order to avoid  type III errors \citep{mosteller:48} consisting of correctly rejecting the null hypothesis for a wrong reason. 
%In some of the studies, the ones described in  \cite{majumder:2013} very specific questions were asked, because the experiments were being conducted to compare results from the lineup protocol with those of conventional tests. 
In those experiments, the structure in the data is strictly controlled in the simulation process, which allows for specific questions to be asked. 
In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is the difference between the two heights. 
Most observers based their decision on this -- besides  `asymmetry' between the towers (a reason suggested by us), most free-form responses for the reason of choice describe a comparison between the red and blue towers. 
In the responses for this lineup we see something that is very typical of responses in lineups: it is only a few panels from the lineup that observers pick, while most panels do not get picked or get picked only a few times. Out of a total of 72~responses, 23 observers picked panel~\#8, 12 observers picked the (data) panel \#13, 10 observers picked panel~\#4, 6 observers picked panel \#7, and all of the remaining panels were picked at most three times. From the panels picked, we see that the top three choices  are all extreme with respect to heights of blue versus red towers:  both panels \#8 and \#13 show a large difference between the height of the towers, while panel \#4 is the only panel, in which the red tower is higher than the blue one. 
% results from the turk study:
% panel:   1 10 11 12 13 14 15 16 18 19  2 20  3   4  5  6  7  8  9 
% #picks:  2  1    0  1 12    3  3    1  1   1  2   1  0 10  2  2  6 23  1

@@

We have conducted a series of experiments (see Table~\ref{tbl:visual_stat} for an overview)
%There have been ten experiments conducted 
using Amazon's Mechanical Turk~\citep{turk} (MTurk) for a variety to purposes: validating the protocol against existing tests, comparing plot designs, and evaluating structure in data analysis problems. In all of the experiments we collected demographic details, such as gender, age bracket, and educational background.  Mining this data will provide a basis for evaluating
%are being used to evaluate 
the effects of different covariates on an individual's ability to evaluate lineups. The design of some of the experimets allow a further investigation into aspects beyond participants' demographics: experiments~5, 6 and 7 can be used to investigate short term learning trends, while experiment~9 allows insight on whether the  positioning of the  data plot in the lineup has an effect on detection rate. 
 %Each of these experiments collected demographic details of the subjects. 
Section~\ref{sec:factor_performance} discusses human factors that we expect to be influential on the performance of the observer. Section~\ref{sec:exp_design} describes the methods used to assess the effects, and Section~\ref{sec:result_socio} summarizes the results of what we have learned about human factors affecting lineup evaluation.

@@

%\subsection{Signal in the Data}
\item{\bf Signal in the Data:}
A visual test statistic is chosen for a particular task. If it is well-designed  departures from the null hypothesis will be visible in the plot, especially if the signal is strong. An important factor that helps an observer to identify the actual data plot in a lineup is the strength of the signal in the data. Simulated data was generated for all of the experiments, except exps.~4 and~9, enabling the study of signal strength and actual data plot detection. In experiment~4 we took a different approach by varying the sampling size, which also allowed us to investigate the effect a design had on detection rate while varying signal/noise ratio. Primarily the results were as expected, that as the signal strength increased observers more frequently picked the actual data plot as different from the null plots.

@@

\begin{enumerate} \leftmargin 5cm  \itemsep 0in
\item {\it Age group}, with categories set to be 18-25, 26-30, 31-35, 36-40, 41-45, 46-50, 51-55, 56-60, above 60.
\item {\it Gender}, male or female.
\item {\it Education level}, with levels being high school or less, some undergraduate courses, undergraduate degree, some graduate courses, and graduate degree.
\item {\it Geographical location}, collected from the IP address of the participants' computer, as latitude, longitude, city and country. 
\end{enumerate}
%
%Instead of collecting exact age, 9 levels of age were collected. They are 18-25, 26-30, 31-35, 36-40, 41-45, 46-50, 51-55, 56-60, above 60. There were five levels of academic information. They are (1) High school or less , (2) Some under grad course, (3) Under graduate degree, (4) Some graduate courses,  (5) Graduate degree.  Geographical locations were collected using the true public ip addresses of the participants' computer. This provides latitude and longitude of the ip address as well as the city and country information. 
%
Let $Y_{ij}$ denote the response from observer $i$ on a lineup $j$, with $Y_{ij} =1$ if the actual data plot is chosen, otherwise $Y_{ij} =0$. The factors are examined in association with the observer's response using a logistic regression with random effect terms:
\begin{equation} \label{eqn:demographic_response}
g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
\end{equation}
where $\pi_{ij}=  E(Y_{ij})$ is the probability that  observer $i$ picks the actual data plot from lineup $j$, %We model this in a generalized mixed effects model of the form
%\begin{equation} \label{eqn:demographic_response}
%g( \pi_{ij} )= \mu + \alpha_{k(i)} + \gamma_{l(i)} + \tau_{m(i)}+ \kappa_{s(i)} + \ell_j,  
%\end{equation}
 $\mu$ is an overall population average, %probability for picking out the data plot from a lineup HH: mu is not a probability, it is on the scale of the link function
 $\alpha$, $\gamma$, $\tau$ and  $\kappa$ are the effects of age group $k(i)$, gender $l(i)$, education level $m(i)$ and country name $s(i)$, respectively, for observer $i$. The term $\ell_j$ is a random intercept predicting lineup difficulty level and we assume independence and normality of the errors, i.e.\ $\ell_j \sim N(0, \sigma_\ell^2)$.  $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$.

@@

It should be noted that we are examining the observer's performance, when we model response as detected or not, but this is not the goal of visual inference. Visual inference is constructed to measure the significance of structure discovered in data. It is expected that some observers will be more skilled at reading data plots, and hence, more readily detect the plot that is different. It is also expected that as observers gain experience in evaluating lineups that they become more proficient in reading data plots, particularly if feedback is given on whether the actual data plot was chosen or not. Choosing the actual data plot will be more difficult in some lineups than others, and indeed should happen purely by chance in some lineups. %, so ``correct'' is more of a convenient rather than ideal word to use for this analysis. 
So in this context, detected, or not, is used as a response to examine individual differences.

@@

%Multiple lineups were sequentially shown to each observer for evaluation. 
Let $Y_{ijk}$ denote the response from observer $i$ on lineup $j$ at their $k$th evaluation attempt, where $Y_{ijk}=1$ if the observer detected the actual data plot otherwise  $Y_{ijk}=0$. Let $\pi_{ijk}=  E(Y_{ijk})$ be the probability that  observer $i$ picks the actual data plot from lineup $j$ in their $k$th attempt. Learning trend is assessed using a generalized mixed effects model of the form
\begin{equation} \label{eqn:trend_response}
g( \pi_{ijk} )= \mu + \alpha_k + u_i +  a_{i} k + \ell_j,  
\end{equation}
where $\mu$ is an overall population average, %probability for picking out the data plot from a lineup, 
$\alpha_k$ is the effect of the $k$th attempt on the probability, using the first attempt as reference, $\alpha_1 = 0$, and $k = 1, ..., K$, $u_i$ and $a_i$ are observer specific random effects, $i = 1, ..., I$. The term, $u_i$ is a random intercept, describing a basic subject-specific ability, with $u_i \sim N(0, \sigma_u^2)$. 
The term $a_i$ is a random slope capturing an individual's specific learning effect over the course of $K$ attempts, where $a_i \sim N(0, \sigma_a^2)$. 
For $\ell_j$ a normal distribution, $N(0, \sigma_\ell^2)$, is assumed, and $\ell_j$ is a random intercept predicting lineup difficulty level. $g(.)$ denotes the {\it logit} link function $g(\pi)=\log(\pi) - \log(1-\pi); 0 \le \pi \le 1$. The inverse link function, $g^{-1}(.)$, from equation~\ref{eqn:trend_response} leads to the estimate of the subject and the lineup specific probability of successful evaluation in the $k$th attempt by a single observer as 
\begin{equation} \label{eqn:trend_power}
\hat p_{ijk} =  g^{-1}(\hat{\mu} + \hat{\alpha}_k + \hat{u}_i +  \hat{a}_i k + \hat{\ell}_j).
\end{equation}

When time taken to evaluate a lineup is used as the response, let $Z_{ijk}$ denote the logarithm of time taken for an observer $i$ to evaluate a  lineup $j$ in his/her $k$th attempt. Let $\mu_{ijk}=  E(Z_{ijk})$ be the average of the (log) of time taken by  observer $i$ to choosing a  panel from lineup $j$ in his/her $k$th attempt. We evaluate this in a mixed effects model of the form
%\hh{
\begin{equation} \label{eqn:trend_time}
Z_{ijk} = \mu + \alpha_1 + \alpha k + u_i +  a_{i} k + \ell_j + \epsilon_{ijk},  
\end{equation}
where $\mu$ represents overall average of log time taken by an observer to evaluate a lineup, $\alpha$ is the average change in log time taken for each additional attempt,  $\alpha_1$ is an offset in log time taken for the first attempt. All other effects are random effects: $u_i$ is a subject-specific intercept representing individual speed of an observer with $u_i \sim N(0, \sigma_u^2)$, $a_i$ is a subject-specific slope representing the deviation of the speed-up (or -down) by attempt $k$, with $a_i \sim N(0, \sigma_a^2)$, $\ell_j$ is a lineup-specific random effect for the time needed to evaluate a lineup, $\ell_j \sim N(0, \sigma_\ell^2)$ and the overall error $\epsilon_{ijk} \sim N(0, \sigma^2)$.
Equation~\ref{eqn:trend_time} leads to the estimate of the subject and the lineup specific time taken for an evaluation in $k$th attempt by a single observer as 
\begin{equation} \label{eqn:trend_time_est}
\hat \mu_{ijk} =  \hat{\mu} + \hat{\alpha_1}+ \hat{\alpha}k + \hat{u}_i +  \hat{a}_i k + \hat{\ell}_j.
\end{equation}

To fit all these mixed effect models the function {\tt lmer()} is used from R package {\tt lme4} by \cite{lme4:2015, lme4:paper}. We employ a normal approximation to obtain $p$-values corresponding to fixed effect parameters estimates. % normal approximation is used to compute $Z$ scores  as the ratio of estimates to  estimated standard errors.
